Joint training of classification model and similarity model for low-resource text classification in chatbot

### 2020-09-10 chatbot_classification/aaai_2021/paper.tex
fix typo

  Task-specific  conversational  chatbot  \cite{wen2016network}  has been applied
  into  many practical products. For example, it can share the working burden of
  human  customer  service  agents  responsible  for  customers' questions about
  certain  products  or services. Another example is the speech-based control in
  devices  like  intelligent  speakers, such as Siri, Google home, Baidu Xiaodu,
  and smart home equipments. Regardless of whether these conversations belong to
  single-turn or multi-turn, an essential technique underlying the chatbot is to
  identify  the  real  intention  behind  a  user's  question or reply. Then the
  chatbot  is caple of conducting expected actions, text reply or accomplish some
  operations.

  It  is  a  challenging task due to some inherent properties. First,
  a customer's  utterance  within  a conversation is usually quite short. Since it
  does  not  contain  enough  information,  short  text \cite{song2014short} are
  thought  to  be  more  ambiguous  compared to long texts such as paragraphs or
  documents,  poseing  a  great challenge \cite{chen2019deep} for classification
  task   \cite{phan2008learning,yan2009dynamic,hua2015short}.   Second,  in  the
  initial  stage  of  building  a  chatbot  , it is often rather hard to collect
  sufficient  labeled  data  to  train  a good model. Because developers have to
  examine tons of real system logs for typical user's utterances, and then label
  them  with expected intentions. Therefore, the key to building a task-specific
  chatbot  with  high  performance  becomes  solving the challenge of short-text
  classification   \cite{sriram2010short}   problem   under   few-shot   setting
  \cite{yu2018diverse}.

@@

  The  second  one,  \emph{similarity model} is also called sentence-pair model.
  Its  motivation  originates  from  the idea of Information Retrival (IR) based
  chatbot   \cite{jafarpour2010filter,   leuski2011npceditor}.   Having   a  Q-A
  (Question-Answer)  pairs dataset and user query Q, the IR based conversational
  system will look up in the Q-A dataset for the pair (Q', A') that best matches
  query  Q  through  semantic  analysis  and  returns  A'  as  the  answer  to Q
  \cite{mnasri2019recent}. In this way, sentence-pair model pre-trained on large
  corpus  of  semantic  similarity  identification task can be applied for being
  used  as  a  tool  to  identify  the class with highest semantic similarity to
  customer's    query.    A    common    model    structure    of    pre-trained sentence-pair model is based on multiple cross-attention
  mechanism  \cite{barkan2020scalable}. Due to
  the   fact   that   many   experiment   results   have   shown   that  RoBERTa
  \cite{liu2019roberta} is an improved version of BERT \cite{devlin2018bert}, and has achieved amazing
  results  on  both  text  classification and sentence-pairs semantic similarity
  tasks, we choose RoBERTa as an important baseline approach in this paper.

@@

  The  above limitation motivates us to propose a joint system that is capble of
  taking  advantagges  from  both  classification model and similarity model. We
  calle  it  SFC,  short  for  similairty model fused with classification Model,
  shown  in  Fig.  \ref{fig:framework}.  To  better train SFC, we further bring in
  multi-task   learning   \cite{caruana1993multitask,collobert2008unified, liu2019multi}.  Our
  basic  idea  cames  from  two  stages. In the first stage, we use an auxiliary
  model  to  select top-K most possible labels for a input. This model can be an
  elastic search  \cite{divya2013elasticsearch}  or  a  text classification model
  trained on current task-specific chatbot data. In the second stage, we build a
  classification  model  whose  main modules are similarity models. Then, this
  structure  derives two goals to train towards, the classification loss for our
  application,  and the similarity loss for the main modules. When the two stages
  are  independently  optimized,  we call \emph{two-stage SFC}. We further find,
  the  the  quality  of  output  from  the  first  stage  might  limit the final
  performance  of  system,  since  the  candidate  labels for each user query is
  always  fixed  in  the  whole  multi-task  training  process. This observation
  motivates  us  to  further  improve  the  two-stage SFC into a joint training,
  called  \emph{joint  SFC}. In this way, the sentence pairs associated with the
  top-K  classes  also  become dynamic, and the text classification model in the
  first  stage  will also be further optimized to provide better top-K result by
  the final training loss.

@@

  \section{Methodology}
  In  this section, we describe our proposed system, similarity model fused with
  classification  model (SFC). We propose two approaches to implement this idea,
  two-stageg SFC, and its more unified version joint-SFC.

@@

  Suppose  we have a training dataset $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$ of
  $N$  data  points,  in  which $x_{i}$ is the user input query and $y_{i}$ is a
  single  class label from a class label set of $C$ classes in total. We 
  generate            a sentence            pair            dataset
  $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$ from $\mathcal{D}$,   where   $(x,   x')_{j}$  is
  a pair of two history queries \footnote{For simplicty, we use query
  to denote a user's question or a response.}, and $l_{j}\in\{0, 1\}$ denotes the
  similarity label that the ones from the same class label are considered
  similar. The size of $\mathcal{D'}$  would be up to $O(N^2)$ if no any
  sampling strategy is used.  This is quite time-consuming  for the  training
  in stage 2 because $N$ can be several hundreds or thousands even under 
  few shot setting in task-specific chatbot scenario.

@@

  \subsection{Datasets}
  Our experiments aim to study the short text classification in the low-resource
  environment  popular  in  the  task-specific conversional chatbot application.
  Hence we choose and configure set datasets to serve this study. Different from
  the  datasets from other short text classification research, such as semantics
  classification  with  very  limited  semantics labels and relatively longer
  input,   the   datasets  adopted  in  our  experiments  are  typically  having
  comparatively  large  number  of  class labels, ranging from several dozens to
  hundreds,  and  each  class  label  is  associated  with  a handful of labeled
  queries with each  query  being usually  one  sentence. Table \ref{tbe:dataset
  statistic} displays their statistics.

@@

  Regarding  the first two datasets, we conduct 3-fold cross validata experients
  that   70   percent   as  training,  15  percent  as  validation  and  testing
  respectively,  and  report  the  averaged  testing results. Regarding the last
  three datasets, we conduct experiments using a sampling method similar to that
  in   \cite{casanueva2020efficient}   yet  in  a  more  sophisticated  few-shot
  settings.  We  fix  a  test  data for each one, and examine their performances
  using  5,  10,  15,  20,  30,  and 50 samples per class label respectively for
  training.  This  is important, as in practice, a task-specific chatbot usually
  starts  with  merely  a  handful of labeled data available in the early stage.
  Besides,  in  the  active  learning framework, building an effective anxiliary
  system  with limited resource is also quite important to help developers label
  data more efficiently.

@@

  Our SFC systems include two-stage SFC with ablation, trained with two tasks
  sperately and both, and joint SFC with all three tasks.

@@

  In two-stage SFC, in equ. \ref{eq:two-stage_SFC_loss} the weights of two tasks
  are  set  as  $\alpha_S=0.125$ and $\alpha_K=0.875$ in our setting. The reason
  for  giving  task  2 relatively more weights is that according to the ablation
  experiments  we  did  in Section \ref{sec:exp}, task 2 contributes more to the
  overall  performance.  In  joint  SFC,  in  equ.  \ref{eq:joint_SFC_loss}, the
  weights  of  three  tasks  are  set  as  $\alpha_S=0.25$,  $\alpha_K=0.25$ and
  $\alpha_C=0.5$  in  our setting. Overrall, the task weights are less sensitive
  in Joint SFC than in two-stage SFC.

@@

  The above analysises illustrate that suitable fusing the classification model and the 
  similarity model to make them cooperate with each other is helpful to better
  performance.

  \subsubsection*{Does joint SFC improve the classfication in stage 1?}
  We  compare  the  top  1  predition and top 5 candidate labels in stage 1 from
  two-stage SFC and joint SFC in Table \ref{tbe:top1_5_accuracy}. We should note
  that  two-stage  SFC  does not optimize the stage 1 result, which is a RoBERTa
  based classification model.

@@

  LEAM  does not perform as well as SFCs, since the first intuitive reason is it
  does  not use pretrained model as encoding module; and another critical reason
  is  in  task-specific  chat applications, it is common to have many intentions
  that are close to some others with a minor difference, and it is hard and even
  impossible  to  name  each  intention  with  a short clear name. One candidate
  solution  is  setting  a most standard sample as the label of that class. When
  using  non-pretrained  models  base  LEAM,  this is applicable. Yet when using
  pretrained  model  based, as the all labels can be hundreds of thousands, then
  this  can not be accommodated in a poplar 32 G Tesla V100 GPU. Actually, joint
  SFC  can  be  kind  of  understood  as  a  generalization form of LEAM. In the
  scenario when there is no clear class label, the relationship between a sample
  and  a class lable is implicatily encoded as that between a sample and another
  sample  from  the  same  class, and this turns into a sentence pair similarity
  model.

@@

  We  conduct  experiments on 5 different settings of hyperparameters, which are
  the  number of candidate class number $K$ and the number of sentence pairs $P$
  for  each  class,  on  ITG  and Amazon-670K datasets to find the most suitable
  setting  of  these  2 hyperparameters. We control the total number of sentence
  pairs  sampled for each user query into about the same level of amount(50-60).
  Through  analyzing the experimenal results shown in Table \ref{tbe:table3}, we
  find  that the best setting of hyperparameters might be different for datasets
  of  different  data  size.  The  class label number of ITG and Amazon-670k are
  similar  in  our  experimental setting, which is around 250. However the total
  amount of data points of ITG is approximately 2 times larger than Amazon-670k.
  Therefore,  the  setting of $K=5$ and $P=10$ is good enough for ITG since most
  of  the  classes  contains  over 10 data points. However, for Amazon-670k, the
  amount  of  data  sample  for  each  class  is  scarce, thus, we may need more
  candidate  classes to sample enough effective sentence pairs for sentence-pair
  model  layer  of  SFC.  In this way, the setting of $K=10$ and $P=5$ becomes a
  good choice. However, in spite of the effect on evaluation performance brought
  by different settings of hyperparameters, joint SFC still steadily outperforms
  two-stage  SFC  by  0.41  percents  in  overall  average  F1 score  on  ITG and
  Amazon-670k datasets.

>>>>

  Task-specific  conversational  chatbot  \cite{wen2016network}  has been applied
  into  many practical products. For example, it can share the working burden of
  human  customer  service  agents  responsible  for  customers' questions about
  certain  products  or services. Another example is the speech-based control in
  devices  like  intelligent  speakers, such as Siri, Google home, Baidu Xiaodu,
  and smart home equipments. Regardless of whether these conversations belong to
  single-turn or multi-turn, an essential technique underlying the chatbot is to
  identify  the  real  intention  behind  a  user's  question or reply. Then the
  chatbot  is capable of conducting expected actions, text reply or accomplish some
  operations.

  It  is  a  challenging task due to some inherent properties. First,
  a customer's  utterance  within  a conversation is usually quite short. Since it
  does  not  contain  enough  information,  short  text \cite{song2014short} are
  thought  to  be  more  ambiguous  compared to long texts such as paragraphs or
  documents,  posing  a  great challenge \cite{chen2019deep} for classification
  task   \cite{phan2008learning,yan2009dynamic,hua2015short}.   Second,  in  the
  initial  stage  of  building  a  chatbot  , it is often rather hard to collect
  sufficient  labeled  data  to  train  a good model. Because developers have to
  examine tons of real system logs for typical user's utterances, and then label
  them  with expected intentions. Therefore, the key to building a task-specific
  chatbot  with  high  performance  becomes  solving the challenge of short-text
  classification   \cite{sriram2010short}   problem   under   few-shot   setting
  \cite{yu2018diverse}.

@@

  The  second  one,  \emph{similarity model} is also called sentence-pair model.
  Its  motivation  originates  from  the idea of Information Retrieval (IR) based
  chatbot   \cite{jafarpour2010filter,   leuski2011npceditor}.   Having   a  Q-A
  (Question-Answer)  pairs dataset and user query Q, the IR based conversational
  system will look up in the Q-A dataset for the pair (Q', A') that best matches
  query  Q  through  semantic  analysis  and  returns  A'  as  the  answer  to Q
  \cite{mnasri2019recent}. In this way, sentence-pair model pre-trained on large
  corpus  of  semantic  similarity  identification task can be applied for being
  used  as  a  tool  to  identify  the class with highest semantic similarity to
  customer's    query.    A    common    model    structure    of    pre-trained sentence-pair model is based on multiple cross-attention
  mechanism  \cite{barkan2020scalable}. Due to
  the   fact   that   many   experiment   results   have   shown   that  RoBERTa
  \cite{liu2019roberta} is an improved version of BERT \cite{devlin2018bert}, and has achieved amazing
  results  on  both  text  classification and sentence-pairs semantic similarity
  tasks, we choose RoBERTa as an important baseline approach in this paper.

@@

  The  above limitation motivates us to propose a joint system that is capable of
  taking  advantages  from  both  classification model and similarity model. We
  call  it  SFC,  short  for  similairty model fused with classification Model,
  shown  in  Fig.  \ref{fig:framework}.  To  better train SFC, we further bring in
  multi-task   learning   \cite{caruana1993multitask,collobert2008unified, liu2019multi}.  Our
  basic  idea  came  from  two  stages. In the first stage, we use an auxiliary
  model  to  select top-K most possible labels for a input. This model can be an
  elastic search  \cite{divya2013elasticsearch}  or  a  text classification model
  trained on current task-specific chatbot data. In the second stage, we build a
  classification  model  whose  main modules are similarity models. Then, this
  structure  derives two goals to train towards, the classification loss for our
  application,  and the similarity loss for the main modules. When the two stages
  are  independently  optimized,  we call \emph{two-stage SFC}. We further find,
  the  the  quality  of  output  from  the  first  stage  might  limit the final
  performance  of  system,  since  the  candidate  labels for each user query is
  always  fixed  in  the  whole  multi-task  training  process. This observation
  motivates  us  to  further  improve  the  two-stage SFC into a joint training,
  called  \emph{joint  SFC}. In this way, the sentence pairs associated with the
  top-K  classes  also  become dynamic, and the text classification model in the
  first  stage  will also be further optimized to provide better top-K result by
  the final training loss.

@@

  \section{Methodology}
  In  this section, we describe our proposed system, similarity model fused with
  classification  model (SFC). We propose two approaches to implement this idea,
  two-stage SFC, and its more unified version joint-SFC.

@@

  Suppose  we have a training dataset $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$ of
  $N$  data  points,  in  which $x_{i}$ is the user input query and $y_{i}$ is a
  single  class label from a class label set of $C$ classes in total. We 
  generate            a sentence            pair            dataset
  $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$ from $\mathcal{D}$,   where   $(x,   x')_{j}$  is
  a pair of two history queries \footnote{For simplicity, we use query
  to denote a user's question or a response.}, and $l_{j}\in\{0, 1\}$ denotes the
  similarity label that the ones from the same class label are considered
  similar. The size of $\mathcal{D'}$  would be up to $O(N^2)$ if no any
  sampling strategy is used.  This is quite time-consuming  for the  training
  in stage 2 because $N$ can be several hundreds or thousands even under 
  few shot setting in task-specific chatbot scenario.

@@

  \subsection{Datasets}
  Our experiments aim to study the short text classification in the low-resource
  environment  popular  in  the  task-specific conversational chatbot application.
  Hence we choose and configure set datasets to serve this study. Different from
  the  datasets from other short text classification research, such as semantics
  classification  with  very  limited  semantics labels and relatively longer
  input,   the   datasets  adopted  in  our  experiments  are  typically  having
  comparatively  large  number  of  class labels, ranging from several dozens to
  hundreds,  and  each  class  label  is  associated  with  a handful of labeled
  queries with each  query  being usually  one  sentence. Table \ref{tbe:dataset
  statistic} displays their statistics.

@@

  Regarding  the first two datasets, we conduct 3-fold cross validation experiments
  that   70   percent   as  training,  15  percent  as  validation  and  testing
  respectively,  and  report  the  averaged  testing results. Regarding the last
  three datasets, we conduct experiments using a sampling method similar to that
  in   \cite{casanueva2020efficient}   yet  in  a  more  sophisticated  few-shot
  settings.  We  fix  a  test  data for each one, and examine their performances
  using  5,  10,  15,  20,  30,  and 50 samples per class label respectively for
  training.  This  is important, as in practice, a task-specific chatbot usually
  starts  with  merely  a  handful of labeled data available in the early stage.
  Besides,  in  the  active  learning framework, building an effective auxiliary
  system  with limited resource is also quite important to help developers label
  data more efficiently.

@@

  Our SFC systems include two-stage SFC with ablation, trained with two tasks
  separately and both, and joint SFC with all three tasks.

@@

  In two-stage SFC, in equ. \ref{eq:two-stage_SFC_loss} the weights of two tasks
  are  set  as  $\alpha_S=0.125$ and $\alpha_K=0.875$ in our setting. The reason
  for  giving  task  2 relatively more weights is that according to the ablation
  experiments  we  did  in Section \ref{sec:exp}, task 2 contributes more to the
  overall  performance.  In  joint  SFC,  in  equ.  \ref{eq:joint_SFC_loss}, the
  weights  of  three  tasks  are  set  as  $\alpha_S=0.25$,  $\alpha_K=0.25$ and
  $\alpha_C=0.5$  in  our setting. Overall, the task weights are less sensitive
  in Joint SFC than in two-stage SFC.

@@

  The above analysis illustrate that suitable fusing the classification model and the 
  similarity model to make them cooperate with each other is helpful to better
  performance.

  \subsubsection*{Does joint SFC improve the classification in stage 1?}
  We  compare  the  top  1  prediction and top 5 candidate labels in stage 1 from
  two-stage SFC and joint SFC in Table \ref{tbe:top1_5_accuracy}. We should note
  that  two-stage  SFC  does not optimize the stage 1 result, which is a RoBERTa
  based classification model.

@@

  LEAM  does not perform as well as SFCs, since the first intuitive reason is it
  does  not use pretrained model as encoding module; and another critical reason
  is  in  task-specific  chat applications, it is common to have many intentions
  that are close to some others with a minor difference, and it is hard and even
  impossible  to  name  each  intention  with  a short clear name. One candidate
  solution  is  setting  a most standard sample as the label of that class. When
  using  non-pretrained  models  base  LEAM,  this is applicable. Yet when using
  pretrained  model  based, as the all labels can be hundreds of thousands, then
  this  can not be accommodated in a poplar 32 G Tesla V100 GPU. Actually, joint
  SFC  can  be  kind  of  understood  as  a  generalization form of LEAM. In the
  scenario when there is no clear class label, the relationship between a sample
  and  a class label is implicitly encoded as that between a sample and another
  sample  from  the  same  class, and this turns into a sentence pair similarity
  model.

@@

  We  conduct  experiments on 5 different settings of hyperparameters, which are
  the  number of candidate class number $K$ and the number of sentence pairs $P$
  for  each  class,  on  ITG  and Amazon-670K datasets to find the most suitable
  setting  of  these  2 hyperparameters. We control the total number of sentence
  pairs  sampled for each user query into about the same level of amount(50-60).
  Through  analyzing the experimental results shown in Table \ref{tbe:table3}, we
  find  that the best setting of hyperparameters might be different for datasets
  of  different  data  size.  The  class label number of ITG and Amazon-670k are
  similar  in  our  experimental setting, which is around 250. However the total
  amount of data points of ITG is approximately 2 times larger than Amazon-670k.
  Therefore,  the  setting of $K=5$ and $P=10$ is good enough for ITG since most
  of  the  classes  contains  over 10 data points. However, for Amazon-670k, the
  amount  of  data  sample  for  each  class  is  scarce, thus, we may need more
  candidate  classes to sample enough effective sentence pairs for sentence-pair
  model  layer  of  SFC.  In this way, the setting of $K=10$ and $P=5$ becomes a
  good choice. However, in spite of the effect on evaluation performance brought
  by different settings of hyperparameters, joint SFC still steadily outperforms
  two-stage  SFC  by  0.41  percents  in  overall  average  F1 score  on  ITG and
  Amazon-670k datasets.
